Upgrades to the SINQ Cold Neutron Source

### 2015-09-15 paper.tex
first draft done
typos

Two variance reduction methods were used in conjunction to perform the simulations in a reasonable amount of time.  First, mesh-based weight windows were generated based on the response of a point detector located inside the AMOR neutron guide, 75 cm from the entrance.  The position gives the detector approximately a two degree view towards to cold source.  The generated weight windows serve to increase the number of neutrons in regions that make large contributions to the point detector, which in this case is the cold source. In order to not bias simulation, the weight windwos cause neutrons entering the cold source are split into many neutrons with a lower statistical weight.

@@

In conjunction with the variance reduction methods described in the previous section, an approximation to how the neutron geuides transport neutrons was also used.  A patch was developed at ORNL and PSI that implements the McStas guide reflectivity model in MCNPX 2.5.0 and 2.7.0 \cite{mcnp_reflectivity, EK_reflectivity}, but the model is analog and is not fully compatible with DXTRAN spheres.  If a neutron is sampled to \emph{not} reflect on a guide surface, it is transmitted into the next cell.  This means that neutron paths have to be sampled multiple times to account for any reflectivity losses, which slows the simulation down.

Some stability issues were also encountered while using the reflectivity patch in the same simulation as a DXTRAN sphere.  In order to ensure stable runs, the simulation was split into two parts.  First, a simulation was done where a surface source was written at the guide entrance surface.  Any neutrons entering the guide were recorded and terminated.  This source was then used in a second simulation with the guide reflectivity activiated.  This step transported the entering neutrons to the end of the guide where a tally was scored.  Again, this process slowed the simulation down since two MCNP simulations needed to be made for each case.  A guide response approximation was used to eliminate the second run, which simply served to transport the neutrons to the end of the guide.

@@

In order to verify that the reconstruction method works as expected, a comparison was made to an exit spectrum calculated with reflectivity model activated.  As mentioned previously, a surface source was first written at the guide entrance, then this was used as the source for a second run with the guide reflectivity model activated.  During the second run, the entry and exit current was tallied.  The entry current was binned in polar angle according to the divisions in the pre-calculated transfer functions.  The exit spectrum was the approximated by   The results of the comparison are shown in Figure \ref{xfer_bench}.  The green line is the total, integrated neutron spectrum at the guide entrance, the blue line is the integrated exit current that was trnaported by the reflectivity model and tallied at the guide exit (the benchmark), and the red curve is the reconstructed exit spectrum.  The maximum error seen is about 70\% at 0.55 \AA{}, but decreases with energy and is almost always within the statistical error of the simulation.  The average 1-norm value (norm divided by number of bins) of the relative error is 7.8\%, and the average 2-norm value is 0.8\%.

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{graphics/xfer_bench.pdf}
\end{center}
\caption{\label{xfer_bench}The reconstructed guide exit current compared to the exit current calculated with a MCNP simulation with guide reflectivity activated.  The right axis is for the exit spectra and the left axis is for the entrance spectrum.  The lower plot shows the recontruction relative error compared to the reflectivity simulation.  The green envelope is the 1-sigma relative error of the reflectivity simulation.}
\end{figure}

@@

\begin{figure}[h]
\begin{minipage}{14pc}
\begin{center}
\includegraphics[scale=0.4,trim={13cm 6cm 4cm 5cm},clip]{graphics/para_geom.pdf}
\end{center}
\caption{\label{parametric_geom}The geometry varied in the REH parametric study. ``A'' is the hole depth, ``B'' is the opening width, and ``C'' is the hole slope.}
\end{minipage}\hspace{5pc}%
\begin{minipage}{11pc}
\begin{center}
\includegraphics[scale=0.4,trim={11.2cm 8cm 5cm 3cm},clip]{graphics/case077.pdf}
\end{center}
\caption{\label{case077}The best-performing horizontal geometry for the REH.}
\end{minipage} 
\end{figure}

@@

Different types of reflector materials and configurations will also be investigated.  The refector sits behind the cold source (the side opposite the target), and reducing cold neutron losses in it may increase the guide exit current.  Adding a belt of liquid H$_2$ to produce a ``cold spot'' for focusing instruments will also be investigated.  In these studies, the flux reconstruction method will once again be used.

>>>>

Two variance reduction methods were used in conjunction to perform the simulations in a reasonable amount of time.  First, mesh-based weight windows were generated based on the response of a point detector located inside the AMOR neutron guide, 75 cm from the entrance.  The position gives the detector approximately a two degree view towards to cold source.  The generated weight windows serve to increase the number of neutrons in regions that make large contributions to the point detector, which in this case is the cold source. In order to not bias simulation, the weight windows cause neutrons entering the cold source are split into many neutrons with a lower statistical weight.

@@

In conjunction with the variance reduction methods described in the previous section, an approximation to how the neutron guides transport neutrons was also used.  A patch was developed at ORNL and PSI that implements the McStas guide reflectivity model in MCNPX 2.5.0 and 2.7.0 \cite{mcnp_reflectivity, EK_reflectivity}, but the model is analog and is not fully compatible with DXTRAN spheres.  If a neutron is sampled to \emph{not} reflect on a guide surface, it is transmitted into the next cell.  This means that neutron paths have to be sampled multiple times to account for any reflectivity losses, which slows the simulation down.

Some stability issues were also encountered while using the reflectivity patch in the same simulation as a DXTRAN sphere.  In order to ensure stable runs, the simulation was split into two parts.  First, a simulation was done where a surface source was written at the guide entrance surface.  Any neutrons entering the guide were recorded and terminated.  This source was then used in a second simulation with the guide reflectivity activated.  This step transported the entering neutrons to the end of the guide where a tally was scored.  Again, this process slowed the simulation down since two MCNP simulations needed to be made for each case.  A guide response approximation was used to eliminate the second run, which simply served to transport the neutrons to the end of the guide.

@@

In order to verify that the reconstruction method works as expected, a comparison was made to an exit spectrum calculated with reflectivity model activated.  As mentioned previously, a surface source was first written at the guide entrance, then this was used as the source for a second run with the guide reflectivity model activated.  During the second run, the entry and exit current was tallied.  The entry current was binned in polar angle according to the divisions in the pre-calculated transfer functions.  The exit spectrum was the approximated by   The results of the comparison are shown in Figure \ref{xfer_bench}.  The green line is the total, integrated neutron spectrum at the guide entrance, the blue line is the integrated exit current that was transported by the reflectivity model and tallied at the guide exit (the benchmark), and the red curve is the reconstructed exit spectrum.  The maximum error seen is about 70\% at 0.55 \AA{}, but decreases with energy and is almost always within the statistical error of the simulation.  The average 1-norm value (norm divided by number of bins) of the relative error is 7.8\%, and the average 2-norm value is 0.8\%.

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{graphics/xfer_bench.pdf}
\end{center}
\caption{\label{xfer_bench}The reconstructed guide exit current compared to the exit current calculated with a MCNP simulation with guide reflectivity activated.  The right axis is for the exit spectra and the left axis is for the entrance spectrum.  The lower plot shows the reconstruction relative error compared to the reflectivity simulation.  The green envelope is the 1-sigma relative error of the reflectivity simulation.}
\end{figure}

@@

\begin{figure}[h]
\begin{minipage}{14pc}
\begin{center}
\includegraphics[scale=0.4,trim={13cm 6cm 4cm 5cm},clip]{graphics/para_geom.pdf}
\end{center}
\caption{\label{parametric_geom}The geometry varied in the REH parametric study. ``A'' is the hole depth, ``B'' is the opening width, and ``C'' is the hole slope.}
\end{minipage}\hfill%
\begin{minipage}{11pc}
\begin{center}
\includegraphics[scale=0.4,trim={11.2cm 8cm 5cm 3cm},clip]{graphics/case077.pdf}
\end{center}
\caption{\label{case077}The best-performing horizontal geometry for the REH.}
\end{minipage} 
\end{figure}

@@

Different types of reflector materials and configurations will also be investigated.  The reflector sits behind the cold source (the side opposite the target), and reducing cold neutron losses in it may increase the guide exit current.  Adding a belt of liquid H$_2$ to produce a ``cold spot'' for focusing instruments will also be investigated.  In these studies, the flux reconstruction method will once again be used.
